Ionospheric Disturbances in Mexican Territory Produced by Objects Entering the Atmosphere from Space \tnoteref{tnote1}

### 2022-11-29 paper/methodology.tex
Update ELSEVIER paper (which compiles) to overleaf at 28-November-2022

\subsection{Meteors Databases}
\label{ssec:databases}
We selected a sample of meteors which were observed in mexican territory from the Geostationary Lightning Mapper \citep{GOODMAN:2013}. Orignally this project was designed to detect ligthning activity in earth's athmosphere, but has been proven that also can detect bolides entering the athmosphere. The detection comes from two satellites called GOES-16 and GOES-17 orbiting the earth in geostationary orbits. %Figure  shows the area where GOES-16 and GOES-17 detect activity on a lightning flash rate map; in mexican territory clearly both satellites can make detections. %Create a figure where the area coverage of GOES-16 and GOES-17 is shown.
We used the interactive database available at \url{https://neo-bolide.ndc.nasa.gov/#/}. These data are publicly available and easily downloaded from the same website. For each event we can obtain the recorded trajectory of meteors and the corresponding light curve. THe GLM satellites have an umbral magnitude for detection of -14. At this magnitude, a meteor is considered a bolide, and is expected to be at least decimeter-sized (in diameter) to reach such brightness. In the other hand, too bright meteors will saturate the detectors, and thus, lowering the quality of data. The result of this factors implies that the range in size of the objects in our sample varies in diameter between decimeter to meter size. %The sample was chosen following the next criteria:%selected in such way we chose the most probable objects to be detected by GPS sources in mexican territory and its surroundings: 
Each event also has assigned a confidence ratio, from low confidence to high, depending in how bright is the event itself and if the trajectory recorded by GLM ressembles (or not) a straight line. We chose only events whose confidence ratio is high, in oreder to be sure we chose the brightest objects, and thus, in the diameter size of bolides, we favored the meter-sized ones. In table \ref{tab:table-meteors} we list the object we chose to do this work, order in chronological order. The columns of the table, from left to right are and ID to enumerate the meteors in the sample, the date and time each meteor was detected, the duration of the detection, their respective coordinates and the estimated height of the meteor over the ground at the time of the detection. GOES-16 and GOES-17 systematically detect the meteors at slightly different positions and at slightly different times, so we calculated the mean of the duration, latitude and longitude reported by both satellites for each event, and used the standard deviation as the uncertainties.

From table \ref{tab:table-meteors} is also clear that the duration of all the bolides detection last less than a second. This obsevation suggests that the bolides remain undetected by the GLM satellites until they get fragmented due to stagnation presure when they release a huge amount of energy and thus they become detectable.

%From the light curves we can estimate the total radiated energy and then convert it to the total kinetic energy through a relation. With this energy we may estimate physical parameters with the cloud fragmentation model form wheeler et al.

%\begin{itemize}
%    \item The objects were detected inside mexican territory and its surroundings.
%    \item The objects were detected by both satellites GOES 16 and GOES 17 (stereo)
%    \item The detection has been assigned a high confidence ratio.
%\end{itemize}

\begin{table*}
  \centering
  \footnotesize
  \caption{List of bolides detected in mexican territory (plus one detected near Venezuela and one detected near Cuba), detected by the Geostationary Lighning mapper. The events are listed in chronological order. The listed duration, latitude and longitude correspond to the mean of the measurements of both GOES satellites. The uncertainties correspond to the respecting mean deviation.}
\label{tab:table-meteors}
\begin{tabular}{rrrrrrrrrr}
\hline
ID & Date of event  & Start Time (UT)  & Duration (seconds) & Latitude (deg) & Longitude (deg)& Altitude (km) & Energy (kT )\\\hline
GLM-00 & 2019-02-01 & 18:17:09 & $2.651\pm 0.4907$ & $22.45 \pm 0.071$ & $-83.50  \pm  0.424$ & 24           & 2.0978196 +/- 0.42361329      \\         
GLM-01 & 2019-05-23 & 16:36:18 & $0.197\pm 0.0000$ & $24.30 \pm 0.000$ & $-101.60 \pm  0.849$ & 28           & 0.011954659 +/- 7.4054404e-3  \\
GLM-02 & 2019-07-18 & 14:30:30 & $0.058\pm 0.0000$ & $27.20 \pm 0.000$ & $-103.15 \pm  0.778$ & 72           & 5.7827709e-3 +/- 3.9334336e-3 \\
GLM-03 & 2019-08-10 & 11:18:48 & $0.199\pm 0.0757$ & $21.50 \pm 0.000$ & $-102.50  \pm 0.849$ & 92           & 0.010556123 +/- 6.6488245e-3  \\
GLM-04 & 2019-10-03 & 07:55:33 & $0.106\pm 0.0297$ & $25.65 \pm 0.071$ & $-96.25 \pm   0.778$ & 74           & 2.9915536e-3 +/- 2.2000998e-3 \\
GLM-05 & 2019-10-09 & 06:08:11 & $0.103\pm 0.0078$ & $23.60 \pm 0.000$ & $-111.95 \pm  0.212$ & 32           & 0.021837042 +/- 0.012429732   \\
GLM-06 & 2019-11-16 & 09:36:04 & $0.396\pm 0.0134$ & $20.30 \pm 0.000$ & $-100.55 \pm  0.919$ & 82           & 7.5423706e-3 +/- 4.9626060e-3 \\
GLM-07 & 2019-11-17 & 15:36:01 & $0.116\pm 0.0035$ & $31.70 \pm 0.000$ & $-117.70 \pm  1.131$ & 88           & 0.022397444 +/- 0.012701445   \\
GLM-08 & 2019-11-19 & 07:57:40 & $0.097\pm 0.1138$ & $20.00 \pm 0.000$ & $-88.40 \pm   1.131$ & 99           & 1.5012507e-3 +/- 1.1909667e-3 \\
GLM-09 & 2019-11-26 & 13:23:20 & $0.078\pm 0.0290$ & $23.90 \pm 0.000$ & $-108.70 \pm  0.849$ & 81           & 4.9551290e-3 +/- 3.4345742e-3 \\
GLM-10 & 2019-12-04 & 09:42:54 & $0.173\pm 0.0028$ & $31.50 \pm 0.000$ & $-113.65 \pm  0.919$ & 77           & 0.029149047 +/- 0.015891049   \\
GLM-11 & 2019-12-15 & 14:50:49 & $0.127\pm 0.0134$ & $27.70 \pm 0.000$ & $-114.10 \pm  0.849$ & 78           & 0.010556123 +/- 6.6488245e-3  \\
GLM-12 & 2019-12-29 & 16:16:35 & $0.062\pm 0.0134$ & $29.60 \pm 0.000$ & $-116.35 \pm  0.919$ & 79           & 4.2084911e-3 +/- 2.9746446e-3 \\
GLM-13 & 2020-01-03 & 14:10:17 & $0.113\pm 0.0085$ & $30.20 \pm 0.000$ & $-117.65 \pm  0.919$ & 74           & 0.011607116 +/- 7.2187549e-3  \\
GLM-14 & 2020-01-06 & 16:39:27 & $0.118\pm 0.0042$ & $31.40 \pm 0.000$ & $-108.20 \pm  0.990$ & 81           & 0.015448801 +/- 9.2392402e-3  \\
GLM-15 & 2020-01-15 & 15:00:33 & $0.213\pm 0.1351$ & $19.45 \pm 0.071$ & $-95.55 \pm   0.919$ & 93           & 0.012559739 +/- 7.7284725e-3  \\
GLM-16 & 2020-02-12 & 09:25:40 & $0.210\pm 0.0226$ & $18.90 \pm 0.000$ & $-93.50 \pm   0.849$ & 90           & 8.2107211e-3 +/- 5.3440405e-3 \\
GLM-17 & 2020-03-03 & 12:33:27 & $0.062\pm 0.0007$ & $18.25 \pm 0.071$ & $-106.35 \pm  0.636$ & 77           & 3.4441157e-3 +/- 2.4922397e-3 \\
GLM-18 & 2020-03-31 & 19:31:52 & $0.105\pm 0.0573$ & $28.45 \pm 0.071$ & $-112.05 \pm  0.636$ & 61           & 7.2469897e-3 +/- 4.7924800e-3 \\
GLM-19 & 2020-04-08 & 16:25:28 & $0.120\pm 0.0926$ & $26.10 \pm 0.000$ & $-93.90 \pm   0.849$ & 78           & 4.0292119e-3 +/- 2.8626279e-3 \\
GLM-20 & 2020-04-18 & 17:43:25 & $0.139\pm 0.0106$ & $29.00 \pm 0.000$ & $-106.55 \pm  0.919$ & 82           & 5.8303967e-3 +/- 3.9618242e-3 \\
GLM-21 & 2020-04-20 & 16:05:22 & $0.318\pm 0.1655$ & $28.15 \pm 0.071$ & $-97.85 \pm   1.061$ & 88           & 0.031378060 +/- 0.016913982   \\
GLM-22 & 2020-04-25 & 11:03:09 & $0.323\pm 0.0813$ & $32.15 \pm 0.071$ & $-111.60 \pm  1.131$ & 84           & 0.021997346 +/- 0.012507576   \\
GLM-23 & 2020-04-28 & 19:31:52 & $0.105\pm 0.0573$ & $28.45 \pm 0.071$ & $-112.05 \pm  0.636$ & 29           & 0.12687833 +/- 0.053736010    \\
GLM-24 & 2020-05-08 & 10:06:16 & $0.490\pm 0.0750$ & $21.60 \pm 0.000$ & $-92.40 \pm   0.849$ & 81           & 0.033207253 +/- 0.017743607   \\
GLM-25 & 2020-07-15 & 19:58:28 & $0.693\pm 0.0495$ & $24.00 \pm 0.000$ & $-108.35 \pm  0.495$ & 53           & 0.020548979 +/- 0.011800641   \\
GLM-26 & 2020-08-07 & 13:29:57 & $0.163\pm 0.0057$ & $28.80 \pm 0.000$ & $-106.05 \pm  0.919$ & 89           & 0.014606987 +/- 8.8040841e-3  \\
GLM-27 & 2020-09-13 & 16:41:59 & $0.184\pm 0.0078$ & $28.45 \pm 0.071$ & $-113.75 \pm  0.919$ & 85           & 0.010995616 +/- 6.8881646e-3  \\
GLM-28 & 2020-09-30 & 12:28:11 & $0.100\pm 0.0078$ & $24.90 \pm 0.000$ & $-110.90 \pm  0.849$ & 83           & 0.014013965 +/- 8.4951275e-3  \\
GLM-29 & 2020-11-16 & 12:28:11 & $0.100\pm 0.0078$ & $24.90 \pm 0.000$ & $-110.90 \pm  0.849$ &106           & 0.052044572 +/- 0.025868548   \\
GLM-30 & 2020-11-17 & 12:53:41 & $0.404\pm 0.0262$ & $23.00 \pm 0.000$ & $-102.45 \pm  0.919$ & 93           & 0.060624247 +/- 0.029367156   \\
GLM-31 & 2020-12-19 & 10:18:14 & $0.407\pm 0.0110$ & $21.95 \pm 0.071$ & $-101.60 \pm  0.990$ & 98           & 0.060272985 +/- 0.029225978   \\
GLM-32 & 2020-12-23 & 09:43:01 & $0.148\pm 0.0014$ & $25.75 \pm 0.071$ & $-111.25 \pm  0.778$ & 81           & 0.012127937 +/- 7.4982028e-3  \\
GLM-33 & 2020-12-29 & 15:20:54 & $0.118\pm 0.0014$ & $16.80 \pm 0.000$ & $-102.20 \pm  0.707$ & 81           & 0.013161054 +/- 8.0470909e-3  \\
GLM-34 & 2021-03-31 & 09:01:17 & $0.753\pm 0.3083$ & $20.15 \pm 0.071$ & $-92.95 \pm  0.212$  & 24           & 0.054259119 +/- 0.026782008   \\
GLM-Ven& 2019-06-22 & 21:25:45 & $4.873\pm 0.0000$ & $14.9 \pm 0.000$  & $-65.8 \pm 0.000$    & 25           & 6.1014359 +/- 0.81239700      \\ \hline
\end{tabular}
\end{table*}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../meteors_map}
  \caption{Positions of events from table \ref{tab:table-meteors} (blue) and table \ref{tab:table-meteors-2}. The events GLM-00/USG-08 actually correpond to the same bolide, but there are little discrepancies about the position where the bolide was detecte. The same applies for the events GLM-23/USG-10 and GLM-Ven/USG-09, which not appears in the map.}
  \label{fig:meteors-map}
\end{figure}

By the other hand, we got another sample of 10 bolidesfrom US Goverment (USG) sensors from the Center for Near Earth Object Studies (CNEOS), publicly available at \url{https://cneos.jpl.nasa.gov/fireballs/}, where we may obtain directly data about bolides position, the date and time each bolide was detected, the energy released at fragmentation, the velocity and the height (the last two not availble for all bolides). As seen in table \ref{tab:table-meteors-2}, the time span is quite larger, and the released energy is generally larger. Both energy distributions are compared directly in figure \ref{fig:boxplot}. Some elements appear in both samples, since are bright enough to be detected regardless the project involved. In USG sample, the total energy of each meteor is obtained directly, but is not the case for the GLM sample. In appendix \ref{app:distance} and \ref{app:energy} we give details about how this total energy is obtained.

\begin{table*}
    \centering
  \footnotesize
  \caption{List of bolides detected in mexican territory (plus one detected near Venezuela and one detected near Cuba), detected by USG sensors.}
\label{tab:table-meteors-2}
\begin{tabular}{rrrrrrrrrrr}
       &               &                  & \multicolumn{4}{c}{Velocity (km/s)}       &                   &                 &\\\cline{4-7}
\multicolumn{1}{c}{ID}&\multicolumn{1}{c}{Date of event}&\multicolumn{1}{c}{Start Time (UT)}&\multicolumn{1}{c}{$v$}&\multicolumn{1}{c}{$v_x$}&\multicolumn{1}{c}{$v_y$}&\multicolumn{1}{c}{$v_z$}&\multicolumn{1}{c}{Latitude (deg)}&\multicolumn{1}{c}{Longitude (deg)}&\multicolumn{1}{c}{Altitude (km)} & \multicolumn{1}{c}{Energy (kT)}\\\hline
USG-01 & 1995-08-05    & 17:14:10         &      &         &           &              & 11.6              &  -104.3         &             &  0.56\\
USG-02 & 1996-07-12    & 14:04:45         &      &         &           &              & 20.7              &   -93.6         &             &  0.11\\ 
USG-03 & 1997-10-09    & 18:47:15         &      &         &           &              & 31.8              &  -106.0         &   37.0      &  0.53\\
USG-04 & 2000-01-18    & 08:33:58         &      &         &           &              & 24.3              &   -94.9         &             &  0.12\\
USG-05 & 2000-08-25    & 01:12:25         &      &         &           &              & 14.5              &  -106.1         &             &   3.1\\
USG-06 & 2005-11-15    & 05:19:07         &      &         &           &              & 26.3              &  -113.4         &   32.4      & 0.089\\
USG-07 & 2015-07-19    & 07:06:26         & 17.8 &   9.4   &  13.0     & 7.8          & 20.6              &   -87.6         &   22.0      & 0.082\\
USG-08 & 2019-02-01    & 18:17:10         & 16.3 &  -2.4   &  13.6     & 8.7          & 22.5              &   -83.8         &   23.7      &   1.4\\
USG-09 & 2019-06-22    & 21:25:48         & 14.9 & -13.4   &   6.0     & 2.5          & 14.9              &   -66.2         &   25.0      &   6.0\\
USG-10 & 2020-04-28    & 05:43:17         &      &         &           &              & 20.1              &  -109.4         &             & 0.076\\\hline  
  \end{tabular}
\end{table*}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{./figures/energies_boxplot}
  \caption{Comparison between released energies of bolides detected by the Geostationary Lightning Mapper and USG sensors.}
  \label{fig:boxplot}
\end{figure}

\subsection{GPS data}
\label{ssec:GPS}
%This material is based on services provided by the GAGE Facility, operated by UNAVCO, Inc., with support from the National Science Foundation and the National Aeronautics and Space Administration under NSF Cooperative Agreement EAR-1724794.

We got RINEX data from 3 to 7 stations depending of the event location and data availability that surround the event place in all directions as possible. A list of the stations where we got RINEX data is available in table%\ref{tab:table-stations}
. Most of the stations lie in mexican territory, but in some cases we required data from other stations to cover events near the mexican frontier at north or south.

%\clearpage
%\onecolumn
%\footnotesize
%\begin{landscape}
%%\begin{table*}
%%    \centering
%  \begin{longtable}{llllllp{12cm}}
%      \caption{List of GPS stations used for this work.}
%      \label{tab:table-stations}
%      \endfirsthead
%      \endhead
%    \hline
%    Station name & Latitude & Longitude & \multicolumn{3}{c}{Events ID}  & Citation \\\hline
%    \multirow{4}{*}{BAR1\hyperlink{Hudnut}{${}^1$}\hyperlink{Hudnut2}{${}^5$}} & \multirow{4}{*}{33.48} & \multirow{4}{*}{-119.03} & \multirow{4}{*}{GLM-07} & \multirow{4}{*}{GLM-12} & \multirow{4}{*}{GLM-13} & UNAVCO Community, Hudnut, Kenneth, King, Nancy, Aspiotes, Aris G., Borsa, Adrian A., Determan, \\
%    &&&&&& Daniel N., Galetzka, John E., Stark, Keith F., 2005, SCIGN-PBO Nucleus GPS Network - BAR1-Santa \\
%    &&&&&& Barbara Island One P.S., The GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5668BHN}.\\\hline
%    \multirow{3}{*}{BLYT\hyperlink{Hudnut}{${}^1$}} & \multirow{3}{*}{33.61} & \multirow{3}{*}{-114.71} &  \multirow{3}{*}{GLM-12} & \multirow{3}{*}{GLM-13}  & & Hudnut, Kenneth, King, Nancy, Aspiotes, Aris G., Borsa, Adrian A., Determan, Daniel N., Galetzka, John \\
%    &&&&&& E., Stark, Keith F., 2006, SCIGN USGS GPS Network - BLYT-Blythe P.S., The GAGE Facility operated \\
%    &&&&&&  by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5HT2MKK}.\\\hline
%    \multirow{3}{*}{CN23} & \multirow{3}{*}{17.26} & \multirow{3}{*}{-88.78} &\multirow{3}{*}{GLM-08} & \multirow{3}{*}{GLM-15} & \multirow{3}{*}{GLM-16} & UNAVCO Community, 2012, COCONet GPS Network - CN23-BelmopanBZCR2012 P.S., The GAGE \\
%    &&&&&& Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5Q23XJH}.\\\hline
%    \multirow{3}{*}{CN25} & \multirow{3}{*}{16.23} & \multirow{3}{*}{-92.13} & \multirow{3}{*}{GLM-15} & & & UNAVCO Community, 2014, COCONet GPS Network - CN25-ComitandDMEX2012 P.S., The GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T57W69G7}.\\\hline
%    \multirow{3}{*}{GCFS} & \multirow{3}{*}{19.31} & \multirow{3}{*}{-81.18} & \multirow{3}{*}{GLM-08} & & & Watts, Anthony, 2016, COCONet GPS Network - GCFS-G\_CAYMAN\_CYM2014 P.S., The GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/7ETV-X536}.\\\hline
%    \multirow{4}{*}{GMPK\hyperlink{Hudnut}{${}^1$}} & \multirow{4}{*}{33.05} & \multirow{4}{*}{-114.83} & \multirow{4}{*}{GLM-10} & & & UNAVCO Community, Hudnut, Kenneth, King, Nancy, Aspiotes, Aris G., Borsa, Adrian A., Determan, \\
%    &&&&&& Daniel N., Galetzka, John E., Stark, Keith F., 2005, SCIGN-PBO Nucleus GPS Network - GMPK-Glamis \\
%    &&&&&& Peak P.S., The GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/WCHN-H687}.\\\hline
%    \multirow{3}{*}{GUAT\hyperlink{Garnier}{${}^2$}} & \multirow{3}{*}{14.59} & \multirow{3}{*}{-90.52} & \multirow{3}{*}{GLM-16} & & & DeMets, Charles, Cosenza-Muralles, Beatriz, 2021, Central America 2018 - Guatemala, The GAGE \\
%    &&&&&& Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/KH2R-K704}.\\\hline
%\multirow{4}{*}{GUAX\hyperlink{Hudnut}{${}^1$}} & \multirow{4}{*}{28.88} & \multirow{4}{*}{-118.29} & GLM-05 & GLM-07 & GLM-11 & Hudnut, Kenneth, King, Nancy, Aspiotes, Aris G., Borsa, Adrian A., Determan, Daniel N., Galetzka, John \\
%    &&&GLM-12& GLM-13& GLM-18 & E., Stark, Keith F., 2001, SCIGN USGS GPS Network - GUAX-Isla Guadalupe P.S., The GAGE Facility \\
%    &&& GLM-25 &GLM-27 &GLM-28& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5GX48T2}.\\
%    &&& GLM-32 &       &     &  \\\hline
%    \multirow{3}{*}{IAGX} & \multirow{3}{*}{29.03} & \multirow{3}{*}{-113.17} & \multirow{3}{*}{GLM-10} & & & Gonzalez-Ortega, Alejandro, Galetzka, John E., Gonzalez, Javier, 2018, CICESE REGNOM GPS Network \\
%    &&&&&& - IAGX-iagxREGNOMmx2018 P.S., The GAGE Facility operated by UNAVCO, Inc., GPS/GNSS  \\
%    &&&&&& Observations Dataset, \url{https://doi.org/10.7283/DGWN-A627}. \\\hline
%    \multirow{2}{*}{INEG} & \multirow{2}{*}{21.85} & \multirow{2}{*}{-102.28} & GLM-25 & GLM-26 & GLM-28  & No citations were found \\
%    &&& GLM-29 & GLM-30 & GLM-31 & \\\hline 
%    \multirow{4}{*}{KVTX} & \multirow{4}{*}{27.55} & \multirow{4}{*}{-97.89} & GLM-01 & GLM-02 & GLM-03  & UNAVCO Community, 2007, PBO GPS Network - KVTX-KingsvilleTX2006 P.S., The GAGE Facility \\
%    &&& GLM-04 & GLM-06 & GLM-19 & operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5J38QH8}.\\
%    &&& GLM-20 & GLM-21 & GLM-24 & \\
%    &&& GLM-26 &&& \\\hline
%    MDO1 & 30.68 & -104.02 & GLM-02 & & & No citations were found\\\hline
%    MGO5 & 30.68 & -104.02 & GLM-21 &GLM-26 &  & No citations were found\\\hline
%    MGW3 & 29.62 & -89.95 & GLM-19 & GLM-21 & GLM-24 & No citations were found\\\hline
%    \multirow{3}{*}{OXTH} & \multirow{3}{*}{16.29} & \multirow{3}{*}{-95.24} & \multirow{3}{*}{GLM-15} & \multirow{3}{*}{GLM-16} & & DeMets, Charles, Cabral-Cano, Enrique, 2008, Oaxaca GPS Network - OXTH-Tehuantepec P.S., The \\
%    &&&&&& GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5Q81B5V}.\\\hline
%    \multirow{3}{*}{OXUM\hyperlink{Graham}{${}^3$}} & \multirow{3}{*}{15.66} & \multirow{3}{*}{-96.50} & \multirow{3}{*}{GLM-34} & & & Cabral-Cano, Enrique, Salazar-Tlaczani, Luis, 2015, TLALOCNet - OXUM-oxum\_tnet\_mx2001 P.S., The \\
%    &&&&&& GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5J964RP}.\\\hline
%    \multirow{2}{*}{P001} & \multirow{2}{*}{31.95} & \multirow{2}{*}{-112.80} & \multirow{2}{*}{GLM-07} & \multirow{2}{*}{GLM-22} & & UNAVCO Community, 2008, PBO GPS Network - P001-Organ\_PipeAZ2007 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5DR2SGP}.\\\hline
%    \multirow{2}{*}{P014} & \multirow{2}{*}{31.97} & \multirow{2}{*}{-11.09} & GLM-10 & GLM-12 & GLM-13 & UNAVCO Community, 2008, PBO GPS Network - P014-Sahuarita\_AZ2007 P.S., The GAGE Facility \\
%    &&& GLM-14 & GLM-22 & & operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5DJ5CMK}.\\\hline
%    \multirow{2}{*}{P807} & \multirow{2}{*}{30.49} & \multirow{2}{*}{-98.82} & GLM-06 & GLM-14 & GLM-21 & UNAVCO Community, 2012, PBO GPS Network - P807-EcRockStPkTX2012 P.S., The GAGE Facility \\
%    &&& GLM-30 &&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5TQ5ZKM}.\\\hline
%    \multirow{2}{*}{PLPX} & \multirow{2}{*}{31.59} & \multirow{2}{*}{-115.15} & \multirow{2}{*}{GLM-10} & & & UNAVCO Community, 2011, PBO GPS Network - PLPX-Las\_PintasMX2010 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5K64G3T}.\\\hline
%    \multirow{2}{*}{PTEX} & \multirow{2}{*}{32.29} & \multirow{2}{*}{-116.52} & GLM-07 & GLM-12 & GLM-13  & UNAVCO Community, 2011, PBO GPS Network - PTEX-Testerazo\_MX2011 P.S., The GAGE Facility \\
%    &&& GLM-27 & GLM-32 & & operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5610XBP}.\\\hline 
%    \multirow{3}{*}{RG06} & \multirow{3}{*}{32.63} & \multirow{3}{*}{-107.86} & \multirow{3}{*}{GLM-22} & & & Sheehan, Anne, 2007, Rio Grande Rift GPS Network - RG06-RG06FaywodNM2006 P.S., The GAGE \\
%    &&&&&& Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5668BFR}.\\\hline
%    \multirow{3}{*}{RG07} & \multirow{3}{*}{32.50} & \multirow{3}{*}{-106.84} & \multirow{3}{*}{GLM-14} & & & Sheehan, Anne, 2007, Rio Grande Rift GPS Network - RG07-RG07CrucesNM2006 P.S., The GAGE  \\
%    &&&&&& Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \\
%    &&&&&& \url{https://doi.org/10.7283/T5KD1W45}. \\\hline
%    \multirow{3}{*}{SG33} & \multirow{3}{*}{31.77} & \multirow{3}{*}{-106.51} & \multirow{3}{*}{GLM-06} & \multirow{3}{*}{GLM-20} & \multirow{3}{*}{GLM-26} & Harder, Steven, Kaip, Galen, Montana, Carlos, 2004, SuomiNet-G GPS Network - SG33-UTEP P.S., The \\
%    &&&&&& GAGE Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset,  \\
%    &&&&&& \url{https://doi.org/10.7283/T50863KQ}. \\\hline
%    \multirow{3}{*}{TGMX}  & \multirow{3}{*}{20.87} & \multirow{3}{*}{-86.87} & \multirow{3}{*}{GLM-34} & & & UNAVCO Community, 2015, COCONet GPS Network - TGMX-PtoMor\_TG\_MX2015 P.S., The GAGE \\
%    &&&&&& Facility operated by UNAVCO, Inc., GPS/GNSS Observations Dataset,  \\
%    &&&&&& \url{https://doi.org/10.7283/T5154FB7}. \\\hline
%    \multirow{2}{*}{TNAM} & \multirow{2}{*}{20.54} & \multirow{2}{*}{-103.97} & GLM-17 & GLM-25 & GLM-28  & UNAVCO Community, 2014, TLALOCNet - TNAM-TNAM\_TNET\_MX2014 P.S., The GAGE Facility \\
%    &&& GLM-29 & GLM-30 & GLM-31 & operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5QF8R4R}.\\\hline
%    \multirow{2}{*}{TNAT} & \multirow{2}{*}{18.13} & \multirow{2}{*}{-98.04} & \multirow{2}{*}{GLM-15} & & & UNAVCO Community, 2014, TLALOCNet - TNAT-TNAT\_TNET\_MX2014 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5G15Z4S}. \\\hline
%    \multirow{2}{*}{TNBA} & \multirow{2}{*}{28.97} & \multirow{2}{*}{-113.55} & GLM-05 & GLM-07 & GLM-09  & UNAVCO Community, 2015, TLALOCNet - TNBA-TNBA\_TNET\_MX2014 P.S., The GAGE Facility  \\
%    &&& GLM-11 & GLM-12 & GLM-13 & operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T57M0688}.\\\hline
%    \multirow{2}{*}{TNCC} & \multirow{2}{*}{18.79} & \multirow{2}{*}{-103.17} & \multirow{2}{*}{GLM-17} & & & UNAVCO Community, 2015, TLALOCNet - TNCC-TNCC\_TNET\_MX2015 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T50R9MSK}.\\\hline
%    \multirow{2}{*}{TNCM} & \multirow{2}{*}{19.50} & \multirow{2}{*}{-105.04} & \multirow{2}{*}{GLM-17} & \multirow{2}{*}{GLM-23} & & UNAVCO Community, 2014, TLALOCNet - TNCM-TNCM\_TNET\_MX2014 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5B856FW}.\\\hline
%    \multirow{2}{*}{TNCN} & \multirow{2}{*}{18.55} & \multirow{2}{*}{-101.97} & \multirow{2}{*}{GLM-29} & \multirow{2}{*}{GLM-33} & & UNAVCO Community, 2016, TLALOCNet - TNCN-TNCN\_TNET\_MX2016 P.S., The GAGE Facility \\
%    &&&&&& operated by UNAVCO, Inc., GPS/GNSS Observations Dataset, \url{https://doi.org/10.7283/T5610XQM}.\\\hline
%    \multirow{4}{*}{TNCU} & \multirow{4}{*}{28.45} & \multirow{4}{*}{-106.79} & GLM-01 & GLM-02 & GLM-03  & UNAVCO Community, 2014, TLALOCNet - TNCU-CuauhtemocTN2014 P.S., The GAGE Facility \\

>>>>

\subsection{Meteor location and trajectory}
\label{ssec:databases}

Data of the Caribbean meteoroid is available from the Geostationary Lightning Mapper (GLM) \citep{GOODMAN:2013} and the Center for Near Earth Object Studies (CNEOS) of the Jet Propultion Laboratory (JPL) of the NASA. We used the interactive database of both projects, available at \url{https://neo-bolide.ndc.nasa.gov/#/} and \url{https://cneos.jpl.nasa.gov/fireballs/}. From these databases, we found basic physical parameters such as event coordinates, altitude, total velocity asn its componentes, total radiated energy and calculated impact energy. These data are shown in table \ref{tab:Meteor-parameters}.

\begin{table*}[t]
    \centering
    \caption{List of meteor basic parameters. Source: \url{https://cneos.jpl.nasa.gov/fireballs/}}
    \begin{tabular}{cc|c}
     \toprule
    Event date  (yyyy-mm-dd)&   & 2019-06-22 \\\hline
    Time (UT) & &21:25:48\\\hline
    Latitude (deg) & & 14.9\\\hline
    Longitude (deg) & & -66.2\\\hline
    Altitude (km) & &25.0\\\hline
    Velocity (km/s) & & 14.9\\\hline
    Duration (seconds)& & 4.873 \\\hline
    \multirow{3}{*}{Velocity components (km/s)}&$v_x$ & -13.4 \\\cline{2-3}
     & $v_y$& 6.0\\\cline{2-3}
     & $v_z$ & 2.5\\\hline
    Total radiated energy (J) & &\SI{294.7e10}{}\\\hline
    Calculated Total Impact energy (kt)&  & 6 \\
    \bottomrule
    \end{tabular}
    \label{tab:Meteor-parameters}
\end{table*}

Row 7 shows the velocity components of the bolide. $v_x$ is the velocity in the equatorial plane, positive towards the prime meridian, $v_z$ is directed towards the celestial north pole and $v_y$ completes the right-handed coordinate system, see figure \ref{fig:ref_velocity} in appendix \ref{app:velocity}.

From GLM data we know that the bolide was visible for 4.873 seconds. The trajectory using the data is shown in figure  \ref{fig:bolide-trajectory}, among with the GLM records.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{../figures/bolide-trajectory}
    \caption{In orange continuous line, the Caribbean bolide estimated trajectory using velocity components from table \ref{tab:Meteor-parameters}. In blue dots, GLM position records.}
    \label{fig:bolide-trajectory}
\end{figure}

  \section{RINEX data}
  \label{sec:GPS-data}
  We collected RINEX data for event from 38 stations that surrounded the impact coordinates for the day of the event. Since the event ocurred near 24:00 UT, we also obtained GPS data for the next day in case of there would be TIDs detected several hours after the fragmentation. We obtained TEC curves using the GOPI software and detrended the resulting time series for each receiver-satellite line of sight. %The spatial distribution of stations is shown in figure \ref{fig:stations_map}, and 
  Table \ref{tab:table-stations} shows the stations names, coordinates and proper citations.

%  \begin{figure}
%      \centering
%      \includegraphics[width=\linewidth]{./figures/caribbean_stations_map}
%      \caption{UNAVCO GPS stations map. The red star shows the positions where the bolide fragmented. The blue arrow shows the direction the bolide was moving in such instant. The green dots represent the positions of the stations we got GPS data.}
%      \label{fig:stations_map}
%  \end{figure}
